Communicative pressures at the semantics-pragmatics interface

### 2016-11-02 paper/paper-v2.tex
draft v2 minor changes

\subsection{Replication \& expressivity}\label{sec:expressivity}
Expressivity has received particular attention from investigations using evolutionary game theory (e.g. \citealt{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}). Under this view, a type's ability to convey and interpret information successfully confers it a higher fitness, a measure that is relative to the success of other types in the population. In the simplest models, fitness directly translates into the proportion of types present in the population after a generational turnover. This association of communicative success within a population with changes in the proportion of types present in it creates a feedback loop that pressures the population towards greater expressivity.

@@

\subsection{Mutation \& learning}\label{sec:learnability}
Iterated learning is a process in which the behavior of one individual serves as learning input for another. This learner, upon acquisition of this behavior, then goes on to produce behavior that serves as input for a new learner. This process can be thought of as a progression through chains of parents and children; the parent produces linguistic data from which the child infers a language. The latter, now a parent, goes on to produce linguistic data for a new generation of learners. Following \citet{griffiths+kalish:2007} we model iterated learning as a repeated process of Bayesian inference in which learners combine the likelihood of a type producing the received learning input with prior inductive biases.

@@

\section{Scalar implicatures}\label{sec:si-case-study}
%
\hl{We assume an integral part of the answer to be that learners are a priori biased towards simpler, more compressed, lexical representations. That is, rational learners should prefer simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}. More narrowly, codifying more semantically is a priori dispreferred and in cases where pragmatics offers conventional means of enrichment, the communicative disadvantage that speakers would otherwise incurr by the use of simpler semantics is leveled. In this way, these domains play a synergic role in overcoming linguistic pressures which may prevent the lexicalization of pragmatic inferences. Crucially, we also show how a different kind of transmission bias in the form of a noisy perception of the environment can lead to the same outcome as that predicted by a learning bias.}

\hl{The mere presence of systematic noise in the transmission of strategies can introduce regularization that looks as if the agents have a learning bias. But, most importantly, this disturbance of transmission fidelity is due to perceptual noise and properties of the environment, not learning biases. In other words, learning biases are clearly not the only transmission biases that can shape evolution alongside functional pressure. Environmental and perceptual noise can play a role too.}

\hl{As touched upon in \S\ref{sec:introduction} we assume learners to be biased towards simpler semantics. That is, learners have a preference for types that can explain the data by codifying less in their lexical representations. More generally, a drive for simplicity has been argued to underpin speaker preferences for brevity and ease of articulation, as well as to pressure languages towards lexical ambiguity and grammatical compression \citep{zipf:1949,grice:1975,piantadosi+etal:2012b, kirby+etal:2015}. As shown by \citet{chater+vitanyi:2003} simplicity not only provides a solution to induction problems by providing a distinction between equally explanatory hypotheses, but has also proven its empirical worth as a longstanding predictive principle in cognitive modelling.}

Scalar implicatures are a particularly well-studied type of conventional pragmatic inference. They are licensed for groups of expressions ordered in terms of informativity, here understood as an entailment induced order. For instance, {\em some} is entailed by {\em all}. If it were true that `All students came to class', it would also be true that `Some students came to class'. However, as already noted in \S\ref{sec:introduction}, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what their use is normally taken to convey. Instead, the use of a less informative expression when a more informative one could have been used can license a defeasible inference that stronger alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that stronger alternatives do not hold since the speaker did not choose them. In this way, `Some students came to class' is strengthened to convey that some but not all students came to class. Conversely, speakers can rely on their interlocutors to draw this inference. The bound that rules out stronger alternatives is thusly not codified in the lexical meaning of weak alternatives but instead pragmatically supplied.

This kind of strengthening corresponds to our previous description of the pragmatic use of lexicon $L_a$. A pragmatic hearer who reasons about a speaker's use of message $m_1$ will associate it more strongly with $s_1$ than with $s_2$ since the latter is already unambiguously associated with $s_2$. Conversely, a pragmatic speaker will reason about her interlocutor's expected interpretation and use the messages at her disposition accordingly.

\begin{centering}
$L_a$ = \bordermatrix{~ & m_1 & m_2 \cr 
                  s_1 & 1 & 0 \cr
                  s_2 & 1 & 1 \cr} \hspace{2cm} $L_b$ = \bordermatrix{~ & m_1 & m_2 \cr 
                  s_1 & 1 & 0 \cr
                  s_2 & 0 & 1 \cr}\\[0.5cm]
\end{centering}

Our initial question can now be narrowed to the case of scalar implicatures by asking for a justification for the lack of lexical upper-bounds in weak scalar alternatives. That is, why semantics such as those of message $m_1$ in $L_a$ are regularly selected for over the alternative of codifying the bound semantically as in $L_b$. More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} and {\em big} were truth-conditionally incompatible with stronger alternatives such as {\em hot}, {\em and}, {\em all} and {\em huge}?  This question is particularly striking considering the number of expressions that license such inferences across languages.

We see two main explanations for the lack of upper-bounds in the lexical meaning of weak scalar expressions. The first is that their truth-conditional compatibility with stronger expressions endows them with a broader range of applicability by allowing them to occur in contexts in which their upper-bounded reading is absent. This can happen when embedded in downward-entailing contexts, when the speaker is likely uncertain about whether the upper bounded reading is true, or when the distinction between an upper-bounded reading and the simple, only lower-bounded reading, is not relevant. For instance, if for all the speaker knows `Some students came' but she doesn't know whether all came, then the use of not upper-bounded {\em some} succinctly conveys her uncertainty. This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then this is preferred over expressing it overtly through longer expressions. For example, by saying {\em some but not all} explicitly. Importantly, although morphosyntactic disambiguation is dispreferred due to its relative length and complexity \citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and override contextual cues that might otherwise mislead the hearer. In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so. Compare this with a hypothetical language that lexicalizes two expressions for each weak scalar expression -- one with and one lacking an upper-bound. We see four conditions along this functionalist explanation that may pressure languages for English-like semantics over this alternative. First, contextual cues are very reliable. Second, morphosyntactic disambiguation is seldom necessary. Third, morphosyntactic disambiguation is only marginally dispreferred. Fourth, larger lexica are costly. Overall, neither condition is convincing as a pivotal explanatory device for such a wide-spread phenomenon. The first two conditions put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of unambiguous communication. It is likely that human language users are very good at retrieving cues from context, but to stipulate that they are so good as to undercut the benefit of safe communication provided by this hypothetical alternative strikes us as too strong of an assumption.  As for the third and fourth condition, these seem mostly like technical solutions without a proper empirical basis.

Instead, considering the regularity and spread of scalar implicatures together with the observation that monomorphemic expressions that lexically rule out stronger alternatives are unattested across languages (\citealt[252-267]{horn:1984},\citealt{horn:1972,traugott:2004,vdAuwera:2010}) suggests that other forces may be at play. In what follows we investigate the hypothesis that the lack of lexicalization of scalar inferences can be accounted for by the relative representational simplicity of lexical meanings lacking an upper-bound over those that explicitly codify it. While we do not want to argue that functional aspects as the ones discussed above do not play a role, we do see a clear benefit in exploring whether matters of learnability would not give us additional explanatory leverage.

Note however that we do not represent the contrast between lexical representations explicitly. Instead, the learning bias towards a lack of upper-bounds in weak scalar alternatives is directly encoded in the learners' prior over types. In principle this difference could be made precise with an adequate representational language, e.g., through measures over representational complexity such as minimal description length.  There is a growing effort to develop such empirically  testable  representational  languages. For  instance, the so-called {\em language of thought} has been put to test in various rational probabilistic models that show encouraging results (see e.g. \citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012a} and references therein). We think that our assumption is well-warranted as a working hypothesis and decide against such an enrichment at present in order to focus on the effects of linguistic pressures predicted by our model instead.

@@

The space of lexica we consider is given in Table \ref{tab:lexica}. As in the preceding examples, lexica are $(2,2)$-Boolean matrices. These are the simplest matrices that allow us to make the contrast between the presence or abscence of an upper-bound and the use of scalar implicatures precise. As implicit in the discussion so far, one may think of $s_1$ as a ``some but not all''-state and of $s_2$ as an ``all''-state. The literal meaning of weak scalar expressions such as English {\em some} then corresponds to a message true of both $s_1$ and $s_2$ in these fragments. Following our assumption for a preference for simple lexical representations, the prior biases learners against lexica in which a message holds true only of state $s_1$. That is, those messages that lexicalize an upper-bound that rules out the ``all''-state $s_2$. All other semantics are assumed to be a priori equally probable. Accordingly, this prior is captured by $P(t_i) \propto n - c \cdot r$, where $n$ is the total number of states and $r$ is the number of messages only true of $s_1$ in $t_i$'s lexicon, $c \in [0,1]$. Increments in the prior parameter $c$ therefore bring about a stronger preference against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.

@@

While there is a total of $16$ possible $(2,2)$-lexica, a number of them are identical both in terms of expressivity and the learning bias. The competition between such types is determined by the initial configuration of a population, i.e., the proportion with which each type starts out. However, this fact can be obscured when averaging across simulations. We therefore focus on this smaller representative subset.\footnote{Simulations conducted with the full space of possible lexica confirm that the results reported here do not hinge on their exclusion.} Lexica $L_1$ to $L_3$ are not optimal for communication because they assign the same state to all their messages. This failure to associate a single form to a meaning both semantically and pragmatically inevitably leads to their disadvantage in language use. $L_4$ and $L_5$ are our target lexica, previously respectively labeled as lexica $L_b$ and $L_a$. They codify upper-bounded semantics for message $m_1$ and a lack thereof. Lastly, $L_6$ is similar to $L_5$ in that two messages are true of the same state but differs from it in assigning upper-bounded semantics to $m_1$.

Combining a signaling behavior with each of these $6$ lexica yields a total of $12$ distinct types. Note in particular  that a type that has conventionalized upper bounds to realize a (quasi-)partition of the relevant semantic space, such as $L_4$, will produce speaker behavior that is {\em almost} indistinguishable from that of a language that lacks upper bounds, but with pragmatic speakers, such as $L_5$. Almost, because there may be slight differences between the probability with which speakers would (erroneously) use a semantically false description and the probability with which speakers would (erroneously) use a pragmatically suboptimal description. Due to this possibly marginal difference between pragmatic $L_4$ and $L_5$, the selection of one type over the other is expected to mainly depend on the learning bias. Things are less clear for literal $L_5$ contrasted with literal/pragmatic $L_4$. The former has a learning advantage but is expected to fare worse in communicative  terms in virtue of ambiguous $m_1$.

@@

%For convenience, the model's parameters are summarized in Table \ref{tab:summary}.
%
%\begin{table}
%\centering
%\begin{tabular}{l|l|l}
%    \multicolumn{1}{c}{parameter} & \multicolumn{1}{c}{explanation} & \multicolumn{1}{c}{locus}\\ \hline
%    $\lambda \geq 1$ & rationality parameter & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\
%    $\alpha \in [0,1)$ & semantics-pragmatics tension & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\ 
%    $|D|$ & number of data produced per parent type & $P(d|t_j)P(t_i|d)$\\
%    $k = |d|$ & number of observations per datum& $P(d|t_j)P(t_i|d)$\\
%    $l \geq 1$ & posterior parameter from sampling to MAP & $P(t_i|d) \propto [P(t_i)P(d|t_i)]^l$\\
%    $c \in [0,1]$ & learning bias for lack of upper-bounds &  $P(t_i)$
%\end{tabular}
%\caption{Summary of model parameters.} 
%\label{tab:summary}
%\end{table}
\subsection{Learning bias}\label{subsec:bias}
According to our hypothesis, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $L_5$-like semantics. It is instructive to first inspect the effect of these pressures in isolation. For this purpose we focus our attention on three pragmatic types.\footnote{Pragmatic reasoning allows language users to refine their (possibly erroneous) choices. Therefore, it is advantageous even for those types that codify an upper-bound lexically.} Pragmatic $L_3$, a type that is lacking in expressivity but is a priori preferred for its lack of upper-bounds. Pragmatic $L_4$, a type that is functionally advantageous but biased against. And pragmatic $L_5$, combining virtues of the latter two.

\paragraph{Expressivity only.} Recall that the replicator dynamics are sensitive to $\lambda$ and $\alpha$ as both have a bearing on a type's fitness. In particular, low $\alpha$ disadvantages types that rely on pragmatic reasoning to the gain of those that codify more semantically. The rationality parameter $\lambda$ has a similar effect for different reasons: Low rationality leads to a less pronounced preference for the choice(s) expected to succeed best in communication. That is, $\lambda$ regulates the strength by which users of $L_5$ associate non-upper-bounded $m_1$ exclusively with the ``some''-state $s_1$ over the ``all''-state $s_2$. Speakers of $L_4$ need not rely on $\lambda$ for this as the association of $m_1$ with $s_1$ is already part of their language's semantics.

The influence of the rationality parameter for $\alpha = 1$ is depicted in Figure \ref{fig:either-R-or-M}.A. As expected, the less expressive $L_3$ speakers fare the worse and are influenced the least by variations in $\lambda$. In contrast, low values of $\lambda$ result in a higher proportion of $L_4$ speakers relative to $L_5$. This is expected given role of rationality in producing more deterministic behavior in users of $L_5$-like languages. As the rationality parameter increases, the functional difference between $L_4$ and $L_5$ is leveled. Overall, the final populations that result only from a pressure towards expressively come close to an even share of pragmatic $L_4, L_5$ and $L_6$ types (the latter follows the same trajectory as $L_5$ in Figure \ref{fig:either-R-or-M}.A).  That is, expressivity alone can not differentiate between these lexica in populations of rational pragmatic language users.

@@

\paragraph{Learnability only.} The effect of iterated learning without a pressure for expressivity is shown in Figure \ref{fig:either-R-or-M}.B for posterior sampling. In line with our expectations, the share of $L_4$ speakers decreases as the bias against upper-bounds increases. In turn, this benefits $L_3$ and, in particular, $L_4$. However, note that even a strong bias against lexical upper-bounds leads only to a moderate advantage for $L_5$ over $L_4$. Furthermore, a pressure only towards learnability promotes functionally defective languages $L_3$.

Inspecting these pressures separately not only gives some intuitions about the parameters' influence, but also highlights some of their broader implications. First and foremost, neither dynamic comes close to converging to a monomorphic population under most parameter configurations. For instance, while $L_4$ speakers can come to take over a substantial proportion of the population, this only happens in a restricted range of low degrees of rationality. Apart from polymorphy, both pressures make undesirable predictions. A pressure only towards expressivity leads to the to the ejection of $L_1$, $L_2$ and $L_3$ from the population. However, it can not explain the regular selection of $L_5$-like semantics over either of its functionally similar alternatives $L_4$ and $L_6$. In contrast, a pressure only towards learnability has a modest but clear effect in differentiating $L_5$ from these alternatives but fails to rule out functionally suboptimal types such as tautological $L_3$.

@@

\paragraph{Expressivity and learnability.} Figure \ref{fig:cost} illustrates the effect of the learning bias for posterior sampling (\ref{fig:cost}.A) and slightly more MAP-like learning (\ref{fig:cost}.B). More detailed results for all types across a sample of $c$-values for $l = 1$ and $l = 3$ are presented in Table \ref{tab:numeric-results}. Overall, these results suggest that in the present setup a weak bias is sufficient to lead to a selection of $L_5$ over $L_4$. As in the simulations that only considered learnability, this effect increases with the bias' strength provided $L_5$ users are pragmatic. Importantly, the addition of a pressure towards expressivity magnifies this effect and dampens the proliferation of functionally suboptimal types advantaged by the learning bias. As stressed above, this suggests that neither the learning bias nor functional pressure alone but their combination may lead to the lack of upper-bounds in the lexical meaning of scalar expressions.

>>>>

\subsection{Expressivity}\label{sec:expressivity}
Expressivity has received particular attention from investigations using evolutionary game theory (e.g. \citealt{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}). Under this view, a type's ability to convey and interpret information successfully confers it a higher fitness, a measure that is relative to the success of other types in the population. In the simplest models, fitness directly translates into the proportion of types present in the population after a generational turnover. This association of communicative success within a population with changes in the proportion of types present in it creates a feedback loop that pressures the population towards greater expressivity.

@@

\subsection{Learnability}\label{sec:learnability}
Iterated learning is a process in which the behavior of one individual serves as learning input for another. This learner, upon acquisition of this behavior, then goes on to produce behavior that serves as input for a new learner. This process can be thought of as a progression through chains of parents and children; the parent produces linguistic data from which the child infers a language. The latter, now a parent, goes on to produce linguistic data for a new generation of learners. Following \citet{griffiths+kalish:2007} we model iterated learning as a repeated process of Bayesian inference in which learners combine the likelihood of a type producing the received learning input with prior inductive biases.

@@

\section{Scalar implicatures}\label{sec:si-case-study}
%
Scalar implicatures are a particularly well-studied type of conventional pragmatic inferences. They are licensed for groups of expressions ordered in terms of informativity, here understood as an entailment induced order. For instance, {\em some} is entailed by {\em all}. If it were true that `All students came to class', it would also be true that `Some students came to class'. However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what their use is normally taken to convey. Instead, the use of a less informative expression when a more informative one could have been used can license a defeasible inference that stronger alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that stronger alternatives do not hold because the speaker used a weaker alternative instead. In this way, `Some students came to class' is strengthened to convey that some but not all students came to class. Conversely, speakers can rely on their interlocutors to draw this inference. A bound that rules out stronger alternatives is thusly not codified in the lexical meaning of weak alternatives but instead pragmatically supplied.

This kind of strengthening is captured by the signaling behavior of pragmatic types introduced in \S\ref{sec:languages+use}: A pragmatic hearer who reasons about a speaker's use of message involving {\em some} will associate it more strongly with a state in which {\em some but not all} holds than with one in which the stronger alternative {\em all} holds, as the {\em all} already unambiguously conveys the latter state. Conversely, a pragmatic speaker will reason about her interlocutor's expected interpretation and use the messages at her disposition accordingly.

Our initial question can now be narrowed to the case of scalar implicatures by asking for a justification for the lack of lexical upper-bounds in weak scalar alternatives. That is, why they are regularly selected for over other alternatives such as that of codifying the bound semantically. More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} and {\em big} were truth-conditionally incompatible with stronger alternatives such as {\em hot}, {\em and}, {\em all} and {\em huge}?  This question is particularly striking considering the number of expressions that license such inferences across languages.

We see two main explanations for the lack of upper-bounds in the lexical meaning of weak scalar expressions. The first is that their truth-conditional compatibility with stronger expressions endows them with a broader range of applicability by allowing them to occur in contexts in which their upper-bounded reading is absent. This can happen when embedded in downward-entailing contexts, when the speaker is likely uncertain about whether the upper bounded reading is true, or when the distinction between an upper-bounded reading and the simple, only lower-bounded reading, is not relevant. For instance, if for all the speaker knows `Some students came' but she doesn't know whether all came, then the use of {\em some} lacking an upper-bound succinctly conveys her uncertainty. This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then this is preferred over expressing it overtly through longer expressions. For example, by saying {\em some but not all} explicitly. Importantly, although morphosyntactic disambiguation may be dispreferred due to its relative length and complexity \citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and override contextual cues that might otherwise mislead the hearer. In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so. Compare this with a hypothetical language that lexicalizes two expressions for each weak scalar expression -- one with and one lacking an upper-bound. We see four conditions along this functionalist explanation that may pressure languages for English-like semantics over this alternative. First, contextual cues are very reliable. Second, morphosyntactic disambiguation is seldom necessary. Third, morphosyntactic disambiguation is only marginally dispreferred. Fourth, larger lexica are costly. Overall, neither condition is convincing as a pivotal explanatory device for such a wide-spread phenomenon. The first two conditions put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of unambiguous communication. It is likely that human language users are very good at retrieving cues from context, but to stipulate that they are so good as to undercut the benefit of safe communication provided by this hypothetical alternative strikes us as too strong of an assumption.  As for the third and fourth condition, these seem mostly like technical solutions without a proper empirical basis.

Instead, the systematicity and typological spread of scalar implicatures together with the observation that monomorphemic expressions that lexically rule out stronger alternatives are unattested across languages (\citealt[252-267]{horn:1984}, \citealt{horn:1972,traugott:2004,vdAuwera:2010}) suggests that other forces may be at play. In what follows we investigate the hypothesis that the lack of lexicalization of scalar inferences may be accounted for by the relative representational simplicity of lexical meanings lacking an upper-bound over those that explicitly codify it. This difference is reflected in a learning bias towards more compressed lexical representations. That is, rational learners should prefer simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}. More narrowly, codifying more semantically is a priori dispreferred and in cases where pragmatics offers conventional means of enrichment, the communicative disadvantage that speakers would otherwise incurr by the use of simpler semantics is leveled. In this way, these domains play a synergic role in overcoming linguistic pressures which may prevent the lexicalization of pragmatic inferences.

While we do not want to argue that functional aspects as the ones discussed above do not play a role, we do see a clear benefit in exploring whether matters of transmission biases would not give us additional explanatory leverage. Note however that we do not represent the contrast between lexical representations explicitly. Instead, the learning bias towards a lack of upper-bounds in weak scalar alternatives is directly encoded in the learners' prior over types. In principle this difference could be made precise with an adequate representational language, e.g., through measures over representational complexity such as minimal description length.  There is a growing effort to develop such empirically  testable  representational  languages. For  instance, the so-called {\em language of thought} has been put to test in various rational probabilistic models that show encouraging results (see e.g. \citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012a} and references therein). We think that our assumption is well-warranted as a working hypothesis and decide against such an enrichment at present in order to focus on the effects of linguistic pressures predicted by our model instead. In line with our aim to analyze the effects of transmission perturbations more broadly, we also show how noisy perception of the environment can lead to the same outcome as that predicted by a learning bias.

@@

The space of lexica we consider is given in Table \ref{tab:lexica}. These $(2,2)$-Boolean matrices are the simplest ones that allow us to make the contrast between the presence or abscence of an upper-bound and the use of scalar implicatures precise. One may think of the state corresponding to the first row of a matrix as a ``some but not all''-state and the second as an ``all''-state. The literal meaning of weak scalar expressions such as English {\em some} then corresponds to a message true of both rows and rows in these fragments. While there are $16$ possible $(2,2)$-lexica, a number of them are identical both in terms of expressivity and the learning bias. The competition between such types is determined by the initial configuration of a population -- the proportion with which each type starts out. However, this fact can be obscured when averaging across simulations. Simulations conducted with the full space of possible lexica confirm that the results reported here do not hinge on their exclusion. We therefore focus on this smaller representative subset. Lexica $L_1$ to $L_3$ are not optimal for communication because they assign the same state to all their messages. This failure to associate a single form to a meaning both semantically and pragmatically inevitably leads to their disadvantage in language use. $L_4$ and $L_5$ are our target lexica. They codify upper-bounded semantics for the message corresponding to the first matrix's column and a lack thereof. Lastly, $L_6$ is similar to $L_5$ in that two messages are true of the same state but differs from it in assigning upper-bounded semantics to the first column's message.

@@

Combining a signaling behavior with each of these $6$ lexica yields a total of $12$ distinct types. Note in particular  that a type that has conventionalized upper bounds to realize a (quasi-)partition of the relevant semantic space, such as $L_4$, will produce speaker behavior that is {\em almost} indistinguishable from that of a language that lacks upper bounds, but with pragmatic speakers, such as $L_5$. Almost, because there may be slight differences between the probability with which speakers would (erroneously) use a semantically false description and the probability with which speakers would (erroneously) use a pragmatically suboptimal description. Due to this possibly marginal difference between pragmatic $L_4$ and $L_5$, the selection of one type over the other is expected to mainly depend on the transmission bias. Things are less clear for literal $L_5$ contrasted with literal/pragmatic $L_4$ as the former has a learning advantage but is expected to fare worse in communicative terms.

@@

%For convenience, the model's parameters are summarized in Table \ref{tab:summary}.
%
%\begin{table}
%\centering
%\begin{tabular}{l|l|l}
%    \multicolumn{1}{c}{parameter} & \multicolumn{1}{c}{explanation} & \multicolumn{1}{c}{locus}\\ \hline
%    $\lambda \geq 1$ & rationality parameter & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\
%    $\alpha \in [0,1)$ & semantics-pragmatics tension & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\ 
%    $|D|$ & number of data produced per parent type & $P(d|t_j)P(t_i|d)$\\
%    $k = |d|$ & number of observations per datum& $P(d|t_j)P(t_i|d)$\\
%    $l \geq 1$ & posterior parameter from sampling to MAP & $P(t_i|d) \propto [P(t_i)P(d|t_i)]^l$\\
%    $c \in [0,1]$ & learning bias for lack of upper-bounds &  $P(t_i)$
%\end{tabular}
%\caption{Summary of model parameters.} 
%\label{tab:summary}
%\end{table}
\subsection{Learning bias}\label{subsec:bias}
Following our assumption of a preference for simple lexical representations, the prior biases learners against lexica in which a message holds true only of the first row. That is, those messages that lexicalize an upper-bound that rules out the ``all''-state. All other semantics are assumed to be a priori equally probable. Accordingly, this prior is captured by $P(t_i) \propto n - c \cdot r$, where $n$ is the total number of states and $r$ is the number of messages only true of row $1$ in $t_i$'s lexicon, $c \in [0,1]$. Increments in the prior parameter $c$ therefore bring about a stronger bias against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.

According to our hypothesis, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $L_5$-like semantics. It is instructive to first inspect the effect of these pressures in isolation. For this purpose we focus our attention on three pragmatic types.\footnote{Pragmatic reasoning allows language users to refine their (possibly erroneous) choices. Therefore, it is advantageous even for those types that codify an upper-bound lexically.} Pragmatic $L_3$, a type that is lacking in expressivity but is a priori preferred for its lack of upper-bounds. Pragmatic $L_4$, a type that is functionally advantageous but biased against. And pragmatic $L_5$, combining virtues of the latter two.

\paragraph{Expressivity only.} The replicator dynamics are sensitive to $\lambda$ and $\alpha$ as both have a bearing on a type's fitness. The influence of the rationality parameter for $\alpha = 1$ is depicted in Figure \ref{fig:either-R-or-M}.A. As expected, the less expressive $L_3$ speakers fare the worse and are influenced the least by variations in $\lambda$. In contrast, low values of $\lambda$ result in a higher proportion of $L_4$ speakers relative to $L_5$. This is expected given role of rationality in producing more deterministic behavior in users of $L_5$-like languages. As the rationality parameter increases, the functional difference between $L_4$ and $L_5$ is leveled. Overall, the outcome from only a pressure towards expressivity come close to an even share of pragmatic $L_4, L_5$ and $L_6$ types (the latter follows the same trajectory as $L_5$ in Figure \ref{fig:either-R-or-M}.A).  This illustrates the problem we started out with: expressivity alone can not differentiate between (near-)functional equivalents to a degree that justifies the systematic prevalence of $L_5$-like semantics.

%In particular, low $\alpha$ disadvantages types that rely on pragmatic reasoning to the gain of those that codify more semantically. The rationality parameter $\lambda$ has a similar effect for different reasons: Low rationality leads to a less pronounced preference for the choice(s) expected to succeed best in communication. That is, $\lambda$ regulates the strength by which users of $L_5$ associate non-upper-bounded $m_1$ exclusively with the ``some''-state $s_1$ over the ``all''-state $s_2$. Speakers of $L_4$ need not rely on $\lambda$ for this as the association of $m_1$ with $s_1$ is already part of their language's semantics.

@@

\paragraph{Learnability only.} The effect of iterated learning with posterior sampling but without a pressure for expressivity is shown in Figure \ref{fig:either-R-or-M}.B. In line with our expectations, the share of $L_4$ speakers decreases as the bias against upper-bounds increases. In turn, this benefits $L_3$ and, in particular, $L_4$. However, note that even a strong bias against lexical upper-bounds leads only to a moderate advantage for $L_5$ over $L_4$. Crucially, a pressure only towards learnability can promote functionally defective languages such as $L_3$.

Inspecting these pressures separately not only gives some intuitions about the parameters' influence, but also highlights some of their broader implications. First and foremost, neither dynamic comes close to converging to a monomorphic population under most parameter configurations. For instance, while $L_4$ speakers can come to take over a substantial proportion of the population, this only happens in a restricted range of low degrees of rationality. Apart from polymorphy, both pressures make undesirable predictions. A pressure only towards expressivity leads to the to the ejection of $L_1$, $L_2$ and $L_3$ from the population but it can not explain the regular selection of $L_5$-like semantics over either of its functionally similar alternatives. In contrast, a pressure only towards learnability has a modest but clear effect in differentiating $L_5$ from these alternatives but fails to rule out functionally suboptimal types such as tautological $L_3$.

@@

\paragraph{Expressivity and learnability.} Figure \ref{fig:cost} illustrates the effect of the learning bias for posterior sampling (\ref{fig:cost}.A) and slightly more MAP-like learning (\ref{fig:cost}.B) when pressured for both expressivity and learnability.\footnote{\hl{Robert suggests to illustrate the effect with a higher l-value. E.g., l = 1 and l = 10}} More detailed results for all types across a sample of $c$-values for $l = 1$ and $l = 3$ are presented in Table \ref{tab:numeric-results}. These results show that a weak bias is sufficient to lead to a selection of $L_5$ over $L_4$ in the present setup. As in the outcomes that only considered learnability, this effect increases with the bias' strength provided $L_5$ users are pragmatic. Importantly, the addition of a pressure towards expressivity magnifies this effect and dampens the proliferation of functionally suboptimal types advantaged by the learning bias. As stressed above, this indicates that neither a learning bias nor functional pressure alone but their combination may lead to the lack of upper-bounds in the lexical meaning of scalar expressions.

\begin{table}
\centering 
\begin{tabular}{l | c c c c c| c c c c c c c c}
\multicolumn{1}{c}{~} & \multicolumn{5}{c}{$l = 1$} & ~ & \multicolumn{5}{c}{$l = 3$}\\ \hline \hline
  c        &  0  & .1  & .5  & .8  & .9 & ~ & 0 & .1 & .5 & .8 & .9\\ \hline \hline
lit. $L_1$ & .03 & .03 & .04 & .04 & .04& ~ & 0 &0 & 0 & 0 & 0\\ 
lit. $L_2$ & .03 & .03 & .02 & .01 & .04& ~ & 0& 0 & 0 & 0 & 0\\
lit. $L_3$ & .03 & .03 & .04 & .04 & .04& ~ & 0& 0 & 0 & 0 & 0\\
lit. $L_4$ & .07 & .07 & .06 & .06 & .05& ~ & 0& 0 & 0 & 0 & 0\\
lit. $L_5$ & .04 & .05 & .05 & .06 & .06& ~ & 0& 0 & 0 & 0 & 0\\
lit. $L_6$ & .04 & .04 & .04 & .04 & .03& ~ & 0& 0 & 0 & 0 & 0\\ \hline
prg. $L_1$ & .03 & .03 & .04 & .04 & .04& ~ & 0& 0 & 0 & 0 & 0 \\
prg. $L_2$ & .03 & .03 & .02 & .01 & .04& ~ & 0& 0 & 0 & 0 & 0 \\
prg. $L_3$ & .03 & .03 & .04 & .04 & .04& ~ & 0& 0 & 0 & 0 & 0 \\ 
prg. $L_4$ & .22 & .22 & .2 & .18 & .17& ~ & .33& .31 & .23 & .15 & .12 \\
prg. $L_5$ & .22 & .23 & .27 & .3  & .32& ~ & .33& .37 & .54 & .7 & .75 \\
prg. $L_6$ & .22 & .22 & .2 & .18 & .17& ~ & .33& .31 & .23 & .15 & .12
\end{tabular}
\caption{Mean proportions of types in $1000$ populations after $20$ generations across bias values $c \in [0,1]$ with $l =1$ and $l = 3$ ($\alpha = 1, \lambda = 30, k = 5$)}%, $\epsilon < 0.005$.}
\label{tab:numeric-results}
\end{table}

The resulting proportion of pragmatic $L_5$ speakers primarily hinges on four parameters. First, the degree to which linguistic behavior is deterministic, controlled by $\lambda$ and $\alpha$, as it plays a role both for expressivity as well as in producing data that allows learners to discriminate this type from others. Second,  the learning bias $c$ which controls the learners preference for simpler lexical representations. Lastly, the posterior parameter $l$, which magnifies the effects of the learning bias in tandem with replication.

As discussed in relation to Figure \ref{fig:cost}.A, posterior sampling can lead to the incumbency of pragmatic $L_5$. However, not even a strong favorable learning bias combined with a pressure for expressivity completely drives out competing types. This is not so for more posterior maximizing behavior. As shown in Figure \ref{fig:prior-posterior}, the range of bias values within which $L_5$ takes over the population increases with MAP-like learning. In other words, the strength of the learning bias required for a given final proportion of $L_5$ speakers strongly depends on learners' inference mechanism. As for the effect of the other parameters not mentioned so far, changes in sequence length influence the population in a predictable way: smaller values lead to more heterogeneous populations that reflect the learner's prior more faithfully. Larger ones lead to more pronounced differences amongst equally preferred types. This is expected insofar as the likelihood that a small sequence was produced by any type is relatively uniform (modulo prior) whereas the likelihood of types with lexica $L_1$ - $L_3$ to produce larger sequences with the same state-message combination is less likely than for pragmatic types using $L_4$ - $L_6$, or literal $L_4$. Thus, while noteworthy, sequence length has no direct bearing on the contrast of interest.
